---
title: "Gresham Lab Floww Cytometry Double Fluorescent Strain Analysis"
author: 'G. Avecilla, N. Brandt, S. Lauer, F. Abdul-Rahman, D. Gresham'
date: '`r Sys.Date()`'
output: html_notebook
---

This notebook contains the code necessary to analysis flow cytometry data in the Gresham Lab. 

To analyze flow cytometry data, you MUST use the latest version of this code, available on the [Gresham Lab github](https://github.com/GreshamLab/flow).

**Experimental overview**

Comparing the signal between double fluorescent strains created from crossing a prACT1::ymCitrine:kanMX near GAP1 strain with either a prACT1::ymCherry:kanMX near MEP2 or prACT1::ymCherry:kanMX between HXT6 and HXT on the Accuri and FACSAria.

*This experient uses the  Accuri flow cytometer and the FACSAria, which is equiped with the following lasers and filters*

Accuri
* Blue laser (488 nm)
* FL1 filter = 514/20nm   GFP
* FL3 filter = 575/25nm   YFP

* Yellow/green laser (552 nm)
* FL2 filter = 610/20nm   mCherry, dtomato
* FL4 filter = 586/15nm   DsRed

FACSAria
* Red laser (633 nm)
* Detector A = 780/60nm   APC-Cy7
* Detector B = 730/45nm   Alexa-700
* Detector C = 660/20nm   APC

* Blue laser (488 nm)
* Detector A = 780/60nm   PE-Cy7
* Detector B = 710/50nm   PerCP-Cy5-5
* Detector C = 610/20nm   PE-Texas Red
* Detector D = 575/25nm   PE, PI, YFP
* Detector E = 530/30nm   FITC, GFP
* Detector F = 488/10nm   SSC

* Violet laser (407 nm)
* Detector A = 525/50nm   AmCyan, BV510
* Detector B = 450/50nm   Pacific Blue, Hoechst, DAPI, BV421

* Yellow/green laser (561 nm)
* Detector A = 610/20nm   mCherry
* Detector B = 586/15nm   DsRed

**Requirements**

In order to run this code you need:

* to predefine your gates using the gating.R script
* the gates.Rdata workspace, which contains the gates to be used in this script
* a .csv sample sheet in with the first XXXX columns set up as in the example below
* user defined variables, see below and chunk 2

**Output**  
This script generates quality control plots in the notebook and a file(s) with a summary of results.
The user can generate the output file(s) in three different styles: 

1. As a dataframe converted from fcs with all or some of the data.
2. As a dataframe with summary statistics (e.g. median FL1 per sample)
3. As a new .fcs file with additional information (e.g. phenotype or sample information) appended

**Libraries**

```{r Libraries, eval=TRUE}
# This is a function that just makes sure you have a package, or installs it for you without prompting

requireInstall <- function(packageName,isBioconductor=F) {
  if ( !try(require(packageName,character.only=T)) ) {
    print(paste0("You don't have ",packageName," accessible, ",
      "I'm gonna install it"))
    if (isBioconductor) {
      source("http://bioconductor.org/biocLite.R")                        
      biocLite(packageName)                                                 
    } else {
      install.packages("packageName", repos = "http://cran.us.r-project.org")
    }
  }
  return(1)
}

#Load libraries
requireInstall("openCyto",isBioconductor=T)
requireInstall("ggcyto",isBioconductor=T)
requireInstall("gridExtra",isBioconductor=T)
requireInstall("tidyverse")

```

  
**Variables and Output Settings**
Variables for the user to set can be found in chunk one. These include both required variables (e.g., working directory), and optional variables (e.g., style of ouput). There are some defaults for these variables.


```{r User Defined Variables}
#working directory
dir = '.'

#file location of fcs and sampleshhet
path.data = paste(getwd(),"/",sep="")

#fcs run sample name
name1 = "Accuri_Run"
name2 = "Aria_Run"

#fcs data to extract
extract.FSC_A <- "Yes"
extract.SSC_A <- "Yes"
extract.FL1_A <- "Yes"
extract.FL2_A <- "Yes"
extract.FL3_A <- "No"
extract.FL4_A <- "No" 
extract.FSC_H <- "No"
extract.SSC_H <- "No"
extract.FL1_H <- "No"
extract.FL2_H <- "No"
extract.FL3_H <- "No"
extract.FL4_H <- "No"
extract.Width <- "No"
extract.Time  <- "No"

#samplesheet parameters
sample.param <- data.frame(SAMPLE=NA, WELL=NA, STRAIN=NA, GENOTYPE=NA, PLOIDY=NA, MEDIA=NA, EXPERIMENT=NA)

#style of output
#FlowSet
save.flowset <- "No"
#filename
folder.flowset <-  paste('DF_Strain_flowdata', Sys.Date(), sep = '_')
#DataFrame - Individual data points
save.df <- "Yes"
file.df <- paste('DF_Strain_df', Sys.Date(), sep = '_')

```


**Read in Data**
*.fcs files must be in a folder with a unique name and have an accompanying .csv samplesheet
The samplesheet name format must be samplesheet_"unique name".csv
It must contain the following columns, in addition you may add additioanl columns depending on your needs
You must also define the parameters under the pData entry

#Example Set of Samplesheet Parameters
Need to define names
* column1 = Well
* column2 = Strain
* column3 = Genotype
* column4 = Ploidy
* column5 = Media
* column6 = Experiment
* column7 = Userdefined


```{r}
#Accuri Data
#Reads in samplesheet
sample.sheet1 <- read.csv(paste(path.data,"samplesheet_",name1,".csv", sep=""))

#reads in FCS files in order outliined in samplesheet
files <- paste(path.data,name1,"/",sort(factor(list.files(paste(path.data,name1,"/", sep=""),full.names=FALSE), levels = paste(sample.sheet1$Well,".fcs",sep="" ), ordered=TRUE)),sep="")
flowData1 <- read.flowSet(files=files, pattern=".fcs", column.pattern="FSC.A|SSC.A|FL1.A|FL2.A", alter.names = TRUE)


#Adds a sample sheet data to the pData of the flowset
#Defines the sample name in the two places needed for the flowframe
#sampleNames must be unique and is used by flowframe to differenate between flowframes
#pData()$name is used as a defualt value in many of the functions used to display FCS data
sampleNames(flowData1) <- paste(gsub(" ","_",sample.sheet1$Strain),"_",sub(" ","_",sample.sheet$Well), sep="")
pData(flowData1)$name <- sampleNames(flowData1)


#Additional samplesheet parameters added to the the FlowSet
pData(flowData1)$Well <- sample.sheet1$Well
pData(flowData1)$Strain <- sample.sheet1$Strain
pData(flowData1)$Genotype <- sample.sheet1$Genotype
pData(flowData1)$Ploidy <- sample.sheet1$Ploidy
pData(flowData1)$Media <- sample.sheet1$Media
pData(flowData1)$Experiment <- sample.sheet1$Experiment


#Aria Data


#Reads in samplesheet
sample.sheet2 <- read.csv(paste(path.data,"samplesheet_",name2,".csv", sep=""))

map <- data.frame(alias = c("FL1.A", "FL2.A"), channels = c("FITC-A", "mCherry-A"))

#reads in FCS files in order outliined in samplesheet
files <- paste(path.data,name2,"/",sort(factor(list.files(paste(path.data,name2,"/", sep=""),full.names=FALSE), levels = paste(sample.sheet2$Well,".fcs",sep="" ), ordered=TRUE)),sep="")
flowData2 <- read.flowSet(files=files, pattern=".fcs", channel_alias = map, alter.names = TRUE)
flowData2 <- read.flowSet(files=files, pattern=".fcs", column.pattern="FSC.A|SSC.A|FL1.A|FL2.A", channel_alias = map, alter.names = TRUE)


#Adds a sample sheet data to the pData of the flowset
#Defines the sample name in the two places needed for the flowframe
#sampleNames must be unique and is used by flowframe to differenate between flowframes
#pData()$name is used as a defualt value in many of the functions used to display FCS data
sampleNames(flowData2) <- paste(gsub(" ","_",sample.sheet2$Strain),"_",sub(" ","_",sample.sheet2$Well), sep="")
pData(flowData2)$name <- sampleNames(flowData2)


#Additional samplesheet parameters added to the the FlowSet
#!!!! REMEMBER THAT YOU NEED TO MAKE SURE THESE MATCH YOUR SAMPLESHEET PARAMETERS
pData(flowData2)$Well <- sample.sheet2$Well
pData(flowData2)$Strain <- sample.sheet2$Strain
pData(flowData2)$Genotype <- sample.sheet2$Genotype
pData(flowData2)$Ploidy <- sample.sheet2$Ploidy
pData(flowData2)$Media <- sample.sheet2$Media
pData(flowData2)$Experiment <- sample.sheet2$Experiment

flowData <- rbind2(flowData1, flowData2)

```


** Base Summary **
Just checks to make sure you collected the amount of data you thought you did, as well as gathering the actual number of counts in each sample and the number of samples to be used in later chunks.

```{r flowSet summaries}
#Check how many cells were counted in each fcs file
total <- fsApply(flowData, each_col, length)[1:length(flowData)] #total counts per sample
print(total)
#Print the medians of data values for each measurement
fsApply(flowData, each_col, median)[1:length(flowData)]

samples.num <- length(flowData)
print(samples.num)
```


** Gating **
Apply all the gates you created to your data set, if you have more then the 4 generic gates, you'll need to create a new segment of gating code
EXAMPLE CODE
START
  GATEDDATA <- Subset(flowData, GATE) 
  GATEDDATACOUNTS <- fsApply(GATEDDATA, each_col, length)[1:samples.num]
  print(GATEDDATACOUNTS)
END

```{r Application of Gates}
##Subset the data by applying sequential gates##

filteredData <- flowData

#apply doublet gate to ALL SAMPLES
#flowData.singlets <- Subset(flowData, pg.singlets) 
#singlets <- fsApply(flowData.singlets, each_col, length)[1:samples.num]
#print(singlets)

#apply debris gates
#filteredData <- Subset(flowData.singlets, pg.nondebris) 
#non.debris <- fsApply(filteredData, each_col, length)[1:samples.num]
#print(non.debris)

```

** Quality control **
#Note that you may need to adjust the axis, the bin size, and the number cols and rows, and the number of fluorescent gates in order to produce the best visulatization for your data

##Gates
```{r eval=FALSE}

#Singlets gate
ggcyto(flowData, aes(x = `FSC.H`, y =  `FSC.A`)) + geom_hex(bins = 512) +  xlim(0,3e6) + ylim(0,3e6) + geom_gate(pg.singlets) + facet_wrap(~name, ncol = 8, nrow = 4) + ggtitle("First flowset - singlets gate")

#Debris gate

ggcyto(flowData, aes(x = `FSC.A`, y =  `SSC.A`)) + geom_hex(bins = 512) +  xlim(0,2e6) + ylim(0,5e5) + geom_gate(pg.nondebris) + facet_wrap(~name, ncol = 8, nrow = 4) + ggtitle("First flowset - nondebris gate")

```



**Data Extraction**
Extracts the data you want based on your selections in the user defined variables
Also remember you need to make sure your samplesheet paramters match with the parameters you wanted in your extracted data file

```{r Data extraction}


#Move filtered data into a dataframe
#Create the empty data frame
filtered.data <- sample.param
if(extract.FSC_A == "Yes"){filtered.data<-cbind(filtered.data, FSC.A=NA)}
if(extract.SSC_A == "Yes"){filtered.data<-cbind(filtered.data, SSC.A=NA)}
if(extract.FL1_A == "Yes"){filtered.data<-cbind(filtered.data, FL1.A=NA)}
if(extract.FL2_A == "Yes"){filtered.data<-cbind(filtered.data, FL2.A=NA)}
if(extract.FL3_A == "Yes"){filtered.data<-cbind(filtered.data, FL3.A=NA)}
if(extract.FL4_A == "Yes"){filtered.data<-cbind(filtered.data, FL4.A=NA)}
if(extract.FSC_H == "Yes"){filtered.data<-cbind(filtered.data, FSC.H=NA)}
if(extract.SSC_H == "Yes"){filtered.data<-cbind(filtered.data, SSC.H=NA)}
if(extract.FL1_H == "Yes"){filtered.data<-cbind(filtered.data, FL1.H=NA)}
if(extract.FL2_H == "Yes"){filtered.data<-cbind(filtered.data, FL2.H=NA)}
if(extract.FL3_H == "Yes"){filtered.data<-cbind(filtered.data, FL3.H=NA)}
if(extract.FL4_H == "Yes"){filtered.data<-cbind(filtered.data, FL4.H=NA)}
if(extract.Width == "Yes"){filtered.data<-cbind(filtered.data, Width=NA)}
if(extract.Time == "Yes"){filtered.data<-cbind(filtered.data, Time=NA)}

#Fill the data frame with the data you want
for(i in 1:length(filteredData)){
  sample <- sampleNames(filteredData)[i]
  well <- as.character(pData(filteredData)$Well[i])
  strain <- as.character(pData(filteredData)$Strain[i])
  genotype <- as.character(pData(filteredData)$Genotype[i])
  ploidy <- as.character(pData(filteredData)$Ploidy[i])
  media <- as.character(pData(filteredData)$Media[i])
  exp <- as.character(pData(filteredData)$Experiment[i])
  if(extract.FSC_A == "Yes"){fsc.a <- exprs(filteredData[[i,1]])}
  if(extract.SSC_A == "Yes"){ssc.a <- exprs(filteredData[[i,2]])}
  if(extract.FL1_A == "Yes"){fl1.a <- exprs(filteredData[[i,3]])}
  if(extract.FL2_A == "Yes"){fl2.a <- exprs(filteredData[[i,4]])}
  if(extract.FL3_A == "Yes"){fl3.a <- exprs(filteredData[[i,5]])}
  if(extract.FL4_A == "Yes"){fl4.a <- exprs(filteredData[[i,6]])}
  if(extract.FSC_H == "Yes"){fsc.h <- exprs(filteredData[[i,7]])}
  if(extract.SSC_H == "Yes"){ssc.h <- exprs(filteredData[[i,8]])}
  if(extract.FL1_H == "Yes"){fl1.h <- exprs(filteredData[[i,9]])}
  if(extract.FL2_H == "Yes"){fl2.h <- exprs(filteredData[[i,10]])}
  if(extract.FL3_H == "Yes"){fl3.h <- exprs(filteredData[[i,11]])}
  if(extract.FL4_H == "Yes"){fl4.h <- exprs(filteredData[[i,12]])}
  if(extract.Width == "Yes"){width <- exprs(filteredData[[i,11]])}
  if(extract.Time == "Yes"){time <- exprs(filteredData[[i,12]])}

  filtered.data <- rbind(filtered.data, cbind(SAMPLE=sample,WELL=well,STRAIN=strain,GENOTYPE=genotype,PLOIDY=ploidy,MEDIA=media,EXPERIMENT=exp,
                    if(extract.FSC_A == "Yes"){FSC.A=fsc.a},
                    if(extract.SSC_A == "Yes"){SSC.A=ssc.a},
                    if(extract.FL1_A == "Yes"){FL1.A=fl1.a},
                    if(extract.FL2_A == "Yes"){FL2.A=fl2.a},
                    if(extract.FL3_A == "Yes"){FL3.A=fl3.a},
                    if(extract.FL4_A == "Yes"){FL4.A=fl4.a},
                    if(extract.FSC_H == "Yes"){FSC.H=fsc.h},
                    if(extract.SSC_H == "Yes"){SSC.H=ssc.h},
                    if(extract.FL1_H == "Yes"){FL1.H=fl1.h},
                    if(extract.FL2_H == "Yes"){FL2.H=fl2.h},
                    if(extract.FL3_H == "Yes"){FL3.H=fl3.h},
                    if(extract.FL4_H == "Yes"){FL4.H=fl4.h},
                    if(extract.Width == "Yes"){Width=width},
                    if(extract.Time == "Yes"){Time=time}))
}  

 #Cleans up DataFrames
  filtered.data<-filtered.data[2:nrow(filtered.data),]
  if(extract.FSC_A == "Yes"){filtered.data$FSC.A<-as.numeric(filtered.data$FSC.A)}
  if(extract.SSC_A == "Yes"){filtered.data$SSC.A<-as.numeric(filtered.data$SSC.A)}
  if(extract.FL1_A == "Yes"){filtered.data$FL1.A<-as.numeric(filtered.data$FL1.A)}
  if(extract.FL2_A == "Yes"){filtered.data$FL2.A<-as.numeric(filtered.data$FL2.A)}
  if(extract.FL3_A == "Yes"){filtered.data$FL3.A<-as.numeric(filtered.data$FL3.A)}
  if(extract.FL4_A == "Yes"){filtered.data$FL4.A<-as.numeric(filtered.data$FL4.A)}
  if(extract.FSC_H == "Yes"){filtered.data$FSC.A<-as.numeric(filtered.data$FSC.A)}
  if(extract.SSC_H == "Yes"){filtered.data$SSC.A<-as.numeric(filtered.data$SSC.A)}
  if(extract.FL1_H == "Yes"){filtered.data$FL1.A<-as.numeric(filtered.data$FL1.A)}
  if(extract.FL2_H == "Yes"){filtered.data$FL2.A<-as.numeric(filtered.data$FL2.A)}
  if(extract.FL3_H == "Yes"){filtered.data$FL3.A<-as.numeric(filtered.data$FL3.A)}
  if(extract.FL4_H == "Yes"){filtered.data$FL4.A<-as.numeric(filtered.data$FL4.A)}
  if(extract.Width == "Yes"){filtered.data$Width<-as.numeric(filtered.data$Width)}
  if(extract.Time == "Yes"){filtered.data$Time<-as.numeric(filtered.data$Time)}
```

***Plots***

```{r}
for(i in 1:length(flowData)){
fvs <- ggplot(flowData[i])+
  geom_hex(aes(x = `FSC.A`, y =  `SSC.A`), bins = 512) + scale_x_log10() + scale_y_log10() + theme_classic() + theme(legend.position="none")
fvf1 <- ggplot(flowData[i])+
  geom_hex(aes(x = `FSC.A`, y =  `FL1.A`), bins = 512) + scale_x_log10() + scale_y_log10() + theme_classic() + theme(legend.position="none")
fvf2 <- ggplot(flowData[i])+
  geom_hex(aes(x = `FSC.A`, y =  `FL2.A`), bins = 512) + scale_x_log10() + scale_y_log10() + theme_classic() + theme(legend.position="none")
f1vf2 <- ggplot(flowData[i])+
  geom_hex(aes(x = `FL1.A`, y =  `FL2.A`), bins = 512) + scale_x_log10() + scale_y_log10() + theme_classic() + theme(legend.position="none")
f1 <- ggplot(flowData[i])+
  geom_density(aes(x = `FL1.A`, y=..count..)) + scale_x_log10() + theme_classic()
f2 <- ggplot(flowData[i])+
  geom_density(aes(x = `FL2.A`, y=..count..)) + scale_x_log10() + theme_classic()

grid.arrange(fvs, f1vf2, fvf1,fvf2, f1, f2, ncol=2, top=as.character(pData(flowData[i])$Genotype))
}


```



** Saving Data **
```{r Saves Data}

#Resave the flowset, this will now include the data from your samplesheet added into the pDATA
#You can also choose to save only gated data by adjusted what flowset you save
#Default is ungated samples
if(save.flowset == "Yes"){write.flowSet(flowData, folder.flowset)}

#The data extracted from the flowset
if(save.df == "Yes"){save(filtered.data, file=paste(file.df,".Rdata",sep=""))}

```


